\section{Conclusion}

We have presented the design of \arti, a novel library that allows one to check
the invariants of a module signature by simulating user interaction with the
module. \arti behaves like a fake client: it calls functions, constructs and
destructs products or sums, and for each element check that the invariants are
verified. The key to performing this in a generic, abstract manner relies on
GADTs, which abstract the different types that may be manipulated into a common
representation.

We identified various performance problems that arise. The library handles them
via a symbolic representation of types in combination with a little bit of
mutable state to avoid handling large, intermediary results in memory.

The result is a self-contained library that wraps the core concepts of
\emph{external testing} and offers clients a cheap and efficient way to test
their programs. The library, for instance, successfully detects infamous issues
such as the AVL re-balancing issue in the standard library of OCaml, with a much
lower cost than a complete Coq proof of the module.

While the library exposes the essence of \emph{external testing} and has already
proven worthwhile, we believe there is potential for improvement and expansion
into a fully-fledged testing library.

% TODO: related work starts here
Bernardy et al.~\cite{DBLP:conf/esop/BernardyJC10} describe a
systematic way of reducing the testing of polymorphic functions to the
testing of specific monomorphic instances of these functions.
%
Given a polymorphic property, the correctness of the reduced
(monomorphic) property entails the correctness of all other
instanciations. This yields a significant reduction in the necessary
test cases.
% They rely on a version of parametricity that does not hold in ML in
% all its generality. For instance, one should not use side effects.
%
They informally argue that their technique is efficient compared to
the standard praxis of substituting \texttt{int} for polymorphic
types.
%
Note however that both solutions to the problem of testing polymorphic
functions must be applied at the meta-level. That is, the user has to
pick the right instanciation of polymorphic type variables; this
cannot be done automatically inside the host language.
